# Spatial Analysis in R
### Lesson Plan for Math 378
__Christensen, Soren A., Lo, Rachel Y., Reed, Brian T.__

__21 April 2015__

### Introduction

Spatial Statistics is a statistical method that uses space directly in the mathematical computations. Spatial Statistics uses distance, area, orientation or any other spacial relationships to analyze, predict, or model datasets where space is involved.

The concept of Spatial Statistics seems to have origins in cartography but throughout the year has contributed to botany (plants), ecology (environment), ethology (animal movement), epidemiology (disease), and economics. Today, it is most commonly used for the analysis of geographic data.

Spatial statistics is unique from analytic methods commonly studied in an undergraduate-level statistics course. The differences can be seen  by the sheer number of difference in fundamental approach, forms of data, and  fields of applicable research.

### Objectives

Introduce this area of statistics to calss, give them exposure.

Be familiar to applciations of the package.

Talk briefly about other packages.

Show how this is applicable to being an analyst in the air force.

### Example

One of the most important application of spatial analysis in wildlife tracking is home range estimation. Many mammals restrict their movements to a home range, rarely venturing far. By estimating an animals home range we can investigate what events effect the home range, and how it changes based on time. 

For example see how these home ranges change from April to May to June.

```{r include =FALSE}
library(dplyr);
library(adehabitatHR);
```

```{r echo = F}
track = read.csv("DeerSample.csv")
track = na.omit(track)

track2 = filter(track, DATETIME >= 40634 & DATETIME <= 40663) #april
track3 = filter(track, DATETIME >= 40664 & DATETIME <= 40694) #may
track4 = filter(track, DATETIME >= 40695 & DATETIME <= 40724) #june

track2S = SpatialPoints(data.frame(track2$LONGITUDE, track2$LATITUDE))
track3S = SpatialPoints(data.frame(track3$LONGITUDE, track3$LATITUDE))
track4S = SpatialPoints(data.frame(track4$LONGITUDE, track4$LATITUDE))

track2K = kernelUD(track2S)
track3K = kernelUD(track3S)
track4K = kernelUD(track4S)

track2H = getverticeshr(track2K)
track3H = getverticeshr(track3K)
track4H = getverticeshr(track4K)

plot(track2H,col = 2)
plot(track3H,col = 3, add = T)
plot(track4H,col = 4, add = T)

legend("topright", title="Months", c("April","May","June"), fill=c(2,3,4), horiz=F)

title(main = "Figure 1. Home ranges for a single female \n deer in 3 different months")
```

The home range varies so much because female deer give birth in late May or early June.The deer is very active in April because she is eating for two, she moves aorund a lot to find sufficient nourishment. This deer most likely gave birth in late may, which is why the home range is smaller, but not as small as in June. In June the deer has given birth to a new fawn, which limits its movements.

Some methods for calculating home ranges calculates them based off a percentile. One such method is the minimum convex polygon. This method creates the smallest possible convex home range which includes all the points in a certain percentile. 

```{r echo = F}
track1 = filter(track, DATETIME >= 40603 & DATETIME <= 40633)
trackSp = SpatialPoints(data.frame(track$LONGITUDE, track$LATITUDE))

trackCp = mcp(trackSp, percent = 99)
track90C = mcp(trackSp, percent = 90)
track80C = mcp(trackSp, percent = 80)
track70C = mcp(trackSp, percent = 70)
track60C = mcp(trackSp, percent = 60)
track50C = mcp(trackSp, percent = 50)

plot(trackCp)
plot(track90C, add = T, col = 2)
plot(track80C, add = T, col = 3)
plot(track70C, add = T, col = 4)
plot(track60C, add = T, col = 5)
plot(track50C, add = T, col = 6)
plot(trackSp, add = T)
```

### Readings

Materials for the students to review before class, and use as references while doing the homework.

An brief article which addresses the importance of spatial analysis in the world. This is important for the students to understand why this topic is being addressed. 

http://www.nytimes.com/2015/01/09/business/international/mapping-the-worlds-problems.html

This is an introductory vignette for basic exercises using adehabitatHR.

http://cran.r-project.org/web/packages/adehabitatHR/vignettes/adehabitatHR.pdf

Main CRAN webpage for adehabtiatHR

http://cran.r-project.org/web/packages/adehabitatHR/index.html

### Homework Questions

Manipulating data with adehabitat

### References

Other references that might be useful

https://sites.google.com/site/rodriguezsanchezf/news/usingrasagis

https://www.movebank.org/node/14620

http://www.rdocumentation.org/packages/adehabitat

http://core.ac.uk/download/pdf/26815643.pdf

http://cran.r-project.org/web/packages/adehabitat/adehabitat.pdf

http://www.mikemeredith.net/blog/1212_Data_for_home_range_analysis_in_R.htm#1

http://cran.r-project.org/web/packages/adehabitatHR/vignettes/adehabitatHR.pdf